Epistemic Semantics in Guarded String Models

### 2020-06-16 tex/draft.tex
Most references are in. Labels for figures.

% Uncomment this line for the final submission
% Or comment it for the submission version with line numbers.
\aclfinalcopy 
%\def\aclpaperid{***} %  Enter the acl Paper ID here

% CHECK THE BIB STYLE FOR THE STARSEM SUBMISSION

@@

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences (McCarthy 1963, Reiter
2001).  The algebraic theory of {\em Kleene Algebra with Tests}
characterizes algebras with elements corresponding to propositions and
event types in our application (Kozen 2003).  The models we propose
are an epistemic extension of guarded string models for KAT, where a
unary operation interpreted as an existential epistemic modality is
included for each agent.  {\em Action models} in dynamic epistemic
semantics introduced the technique of constructing epistemic models
from primitive alternative relations on events, in order to capture
the epistemic consequences of perceptual events (Baltag, Moss, and
Solecki 1998). This is the basis for our construction of epistemic
alternative relations.  Literature on finite state methods in
linguistic semantics has used event strings and sets of event strings
to theorize about tense and aspect in natural language semantics
(Fernando 2003, Carlson 2009) and to express propositions (Fernando
2017).  Literature on finite state intensiononal semantics has 
shown how to do the semantics of intensional complementation
including indirect questions in a setting where propositions are
represented using finite state machines, and compositional semantics
is expressed in a finite state calculus (Rooth 2017, Collard 2018).
We adopt this in our syntax-semantics interface for
English.

@@

\section{Epistemic guarded string models}
Figure 1 shows an Epik program that describes a possible worlds model for two agents
with information about one coin, and events of the agents semi-privately
looking at the coin. The line beginning with {\tt state} lists the basic
stative propositions. To illustrate syntax, a second proposition $t$ (tails) is
included.  The line beginning with {\tt constraint} defines compatibilities among the
propositions: the coin is heads or tails and not both.  The lines beginning with {\tt
  event} declare events, their preconditions, and their effect on state, following
the format in \ex{event1}.  Finally the lines beginning with {\tt agent} define {\em
  event alternative} relations for agents.  Each clause with an arrow has a single
event symbol on the left, and a disjunction of alternative events on the right of the
arrow. The interpretation of Amy's alternatives for $b_1$ (Bill peeks at heads), is
that when $b_1$ happens, for Amy either $b_1$ or $b_0$ (Bill peeks at tails) could
be happening.

Kleene Algebra with Tests is an algebraic theory that is defined by equations and
inequalities, which a has model classes including guarded string models, relational
models, finite models, and matrix models.  This paper focuses on defining a family of
concrete guarded string algebras, the elements of which are sets of guarded strings.
Definitions and notation mostly follow Kozen (2003).  Additional syntax and semantics
is included to model multi-agent epistemic semantics.  Guarded strings over a finite
alphabet $E$ are like ordinary strings, but with truth assignments to a set $T$ of
primitive propositions (primitive tests) alternating with the symbols from $E$.  In
the algebra described by Figure 1, $E$ is the set of events $\{a_1,a_0,b_1,b_0\}$,
and in the elevator example, $\{u,d\}$.  In the elevator example, $T$ is $\{p,q\}$,
and in the coin example it is $\{h,t\}$.  We write truth assigments to the primitive
propositions that observe the constraint as Boolean vectors. In the coin example, we
get the vectors 01, 10 and in the elevator example, the vectors 00, 01, 10, 11.
Guarded strings are strings of events, alternating with such vectors, and starting
and ending with vectors.  \ex{gs1} gives the encoding as guarded strings of the
worlds in \ex{elevator1} and \ex{coin1}.

@@

\caption{Epik program describing a possible-worlds event sequence model for two agents
  with information about one coin, and events of the agents semi-privately looking
at the coin.}
\end{figure}

@@

It remains to define the Kripke relation on guarded strings from an agent specification
as in Figure 1. An an agent specification pairs each bare event with a set of bare events,
and so determines a relation between bare events, call it relation $R_a$ for an
agent $e$.  This determines
a relation $\hat{R_a}$ between decorated events, see \ex{Rhat}.
It is generalized to a relation $\doublehat{R}$ on arbitrary guarded strings by
relational Kleene star, as defined in \ex{Rhathat}. The operation $\cdot$ in the
definition is KAT product, which enforces matching of tests.
The number of terms $n$ is construed as satisfying $n\geq 1$.
{\em What is desired for $n=0$?}

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics (cite Karttunen etc.)  Current implementations allow for the
definition of functions with the status of defined operators on regular sets and
relations.  Such definitions are used to define an embedding of epistemic KAT in a
string algebra. The methodology follows Section 2 closely.  Let ${\cal K}$ be an
epistemic algebra as described in Section 2.  A given element of ${\cal K}$ is
represented in the the string algebra by the very same set of strings, i.e. by a set
of strings that have the form of a sequence of bare event symbols, with interleaved
Boolean vectors. Product in the KAT can not be modeled as concatenation in the string
algebra, because this would not enforce identity of states, and would result in
lengthening Boolean vectors at the concatenation point.  Instead, KAT product and KAT
Kleene star are defined operations in the string algebra, see Figure 2.  The
operations concatenate in the string algebra, delete strings with non-matching tests,
and then delete the second of two tests create a well-formed guarded string.

@@

\caption{Translation into Fst of KAT product and KAT Kleene star.}
\end{figure}

A given bare event such as $a_1$ (Aly looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, semantically $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure 1 as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$, and an algebra of
propositions as regular sets of strings drawn from this space of worlds.

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure 1 is a relation between bare events for each
agent. This determines a relation in the guarded string algebra a relation between
bare events decorated with compatible tests.  For agent Aly, this is the relation
described in \ex{alyrel1} as a set of ordered pairs.

@@

\caption{Definition in Fst of the Kleene concatenation closure of
a relation between guarded strings.}
\end{figure}

@@

This scheme provides for an interpretation of the language of Epik
terms that is defined in Figure 4.  An Epik specifications such as
Figure 1 is translated to a straight-line program of the finite state
calculus that defines constants and functions, including the ones from
Figures 2 and 3.  In Xfst or Foma, which are interpreters for the
finite state calculus, the program is read, and then propositional
terms can be mapped to finite state machines that represent sets of
guarded strings, interpreted as propositions.

@@

\caption{The language of Epik terms. Propositional terms are interpreted
as sets of guarded strings.}
\end{figure}

An architecture of interpretation by transtlation is employed, where English
sentences are mapped to terms in the logical language (--) via an interpreted
grammar, and these terms are in turn interpreted as propositions (sets of possible
worlds). For the latter, there are options of translation into the finite state
calculus in order to represent propositions as finite state machines (Section 3), and
representation in Haskell via lazy lists of guarded strings (Section 4).  The grammar
is a semantically interpreted multi-modal categorial grammar, consisting of a lexicon
of words, their categorial types, and interpretations in a logical lambda language.
\ex{phenomena1} lists that are covered.

\begin{figure}  
\begin{tabular}[t]{l|l}
  Basic statives & It's heads. \\
  & It's tails.\\ \hline
That-complement & Amy knows \\ & \hspace{2mm} that it's heads. \\ \hline
Wh-complement & Amy knows \\ & \hspace{2mm} whether its heads. \\ \hline
Negation & Bob doesn't know \\ & \hspace{2mm} that it isn't heads. \\ \hline 
Tensed and base &  Bob knows \\
verbal forms & \hspace{2mm} that it's heads.      \\ 
\hspace{1mm} & Bob doesn't know \\ & \hspace{2mm} that it's heads.\\ \hline
Sentence conjunction & It's heads and Bob \\ & \hspace{2mm} doesn't know \\ & \hspace{2mm} that it's heads \\ \hline
Predicate conjunction & Bob knows that \\ & Amy knows whether \\ & it's heads and
\\ & doesn't know that \\ & Amy knows that \\ & it's heads.
\end{tabular}
\caption{Phenomena covered in the English grammar fragment.}
\end{figure}

@@

\ex{lexicon1} gives illustrative lexical entries. The grammar and semantics are in
certain way optimized for a simple fragment of English concerned with clausal
complementation.  The agent names {\em Amy} and {\em Bob} contribute the epistemic
alternative relations for those agents, rather than individuals.  This is possible
because the agents are never arguments of extensional predicates, so what matters
about the agents is their epistemic alternative relations.  The root verb {\em know}
contributes existential modal force.  The complementizers {\em that} and {\em
  whether} are the heads of their dominating clauses, and assemble an alternative
relation, modal force, and proposition contributed by the complement.  These
complementizers introduce the dual via two negations, in order to arrive universal
modal force.  These moves are offered here as a way of constructing a compact
interpreted grammar. They can easily be reformulated in a more comprehensive
interpreted grammar of English.

\begin{figure}  
\begin{tabular}[t]{l|c|c}
Amy & $e$ & ${R}_a$\\
Bob & $e$ & ${R}_b$\\
it & $d$ & $d$ \\
heads & $d \backslash_D t$ & $\lambda x.0^c{\cdot}h$ \\
tails & $d \backslash_D t$ & $\lambda x.0^c{\cdot}!h$ \\
is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P.\lambda x.P x$ \\
knows & $(e \backslash t)\slash_M t$ &  $\lambda p.\lambda R. {\Diamond}R p$\\
that  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
whether  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
& \multicolumn{2}{r}{$+ {\sim}(m p R)$} \\
\end{tabular}
\caption{Partial categorial grammar lexicon. The first column has a word form.
  the second column a categorial type, and third column a semantic translation in
a logical language that extends the Epik term language with lambda.}
\end{figure}

@@

The semantic translations in the third column of \ex{lexicon1} use the Epik term
language, incremented with lambda.  The body of $\lambda x.0^c{\cdot}h$, which is the
semantic lexical entry for {\em heads}, is a term denoting the set of all worlds
where the coin is heads, represented as the set of all guarded strings that end with
a Boolean valuation where the primitive proposition $h$ (it's heads) is true.  There
is $\lambda x$ at the front because of a correspondence the grammar formalism uses a
correspondence between syntactic and semantic types. However, it does not bind
anything, because sentences such as {\em it isn't heads} have an expletive subject.
The body ${\Diamond}R p$ of $\lambda p.\lambda R. {\Diamond}R p$, which is the
semantic lexical entry of {\em knows}, is an Epik term denoting the pre-image of the
set of worlds $p$ according to the relation $\doublehat{R}$ between guarded strings
that is determined by the event-level relation $R$.  This is not the right semantics
for {\em Amy knows that it's heads}, because it is an existential modality ${\Diamond_R}p$,
rather than an universal modality ${\Box_R}p$.  This is corrected by the complementizer
{\em that} or {\em whether}, which introduces the dual.

>>>>

% Uncomment this line for the final submission
% Or comment it for the submission version with line numbers.
\aclfinalcopy 
%\def\aclpaperid{***} %  Enter the acl Paper ID here
% Make the references come out black. TAKE THIS OUT for submission.
\definecolor{darkblue}{rgb}{0, 0, 0}
% CHECK THE BIB STYLE FOR THE STARSEM SUBMISSION

@@

The approach taken here synthesizes five antecedents in a certain way.
John McCarthy's {\em Situation Calculus} is the source of the idea of
constructing possible worlds as event sequences
\citep{mccarthy1963situations,reiter2001knowledge}.  The algebraic
theory of {\em Kleene Algebra with Tests} characterizes algebras with
elements corresponding to propositions and event types in our
application \citep{kozen2001automata}.\marginpar{2001 or 2003?}  The
models we propose are an epistemic extension of guarded string models
for KAT, where a unary operation interpreted as an existential
epistemic modality is included for each agent.  {\em Action models} in
dynamic epistemic semantics introduced the technique of constructing
epistemic models from primitive alternative relations on events, in
order to capture the epistemic consequences of perceptual events
\citep{baltag1999logic}. This is the basis for our construction of
epistemic alternative relations.  Literature on finite state methods
in linguistic semantics has used event strings and sets of event
strings to theorize about tense and aspect in natural language
semantics \citep{fernando2004finite,fernando2007observing,carlson2009tense} and to
express intensionality \cite{fernando2017intensions}.  Literature on
finite state intensiononal semantics has discussed how to do the
semantics of intensional complementation including indirect questions
in a setting where compositional semantics is expressed in a finite
state calculus \citep{rooth2017finite,collard2018finite}.  We adopt
this in our syntax-semantics interface for English.

@@

\section{Epistemic guarded string models}
Figure \ref{epikfigure} shows an Epik program that describes a possible worlds model for two agents
with information about one coin, and events of the agents semi-privately
looking at the coin. The line beginning with {\tt state} lists the basic
stative propositions. To illustrate syntax, a second proposition $t$ (tails) is
included.  The line beginning with {\tt constraint} defines compatibilities among the
propositions: the coin is heads or tails and not both.  The lines beginning with {\tt
  event} declare events, their preconditions, and their effect on state, following
the format in \ex{event1}.  Finally the lines beginning with {\tt agent} define {\em
  event alternative} relations for agents.  Each clause with an arrow has a single
event symbol on the left, and a disjunction of alternative events on the right of the
arrow. The interpretation of Amy's alternatives for $b_1$ (Bill peeks at heads), is
that when $b_1$ happens, for Amy either $b_1$ or $b_0$ (Bill peeks at tails) could
be happening.

Kleene Algebra with Tests is an algebraic theory that is defined by
equations and inequalities, which a has model classes including
guarded string models, relational models, finite models, and matrix
models.  This paper focuses on defining a family of concrete guarded
string algebras, the elements of which are sets of guarded strings.
Definitions and notation mostly follow \cite{kozen2001automata}.
Additional syntax and semantics is included to model multi-agent
epistemic semantics.  Guarded strings over a finite alphabet $E$ are
like ordinary strings, but with truth assignments to a set $T$ of
primitive propositions (primitive tests) alternating with the symbols
from $E$.  In the algebra described by Figure \ref{epikfigure}, $E$ is the set of
events $\{a_1,a_0,b_1,b_0\}$, and in the elevator example, $\{u,d\}$.
In the elevator example, $T$ is $\{p,q\}$, and in the coin example it
is $\{h,t\}$.  We write truth assigments to the primitive propositions
that observe the constraint as Boolean vectors. In the coin example,
we get the vectors 01, 10 and in the elevator example, the vectors 00,
01, 10, 11.  Guarded strings are strings of events, alternating with
such vectors, and starting and ending with vectors.  \ex{gs1} gives
the encoding as guarded strings of the worlds in \ex{elevator1} and
\ex{coin1}.

@@

\caption{Epik program describing a possible-worlds event sequence model for two agents
  with information about one coin, and events of the agents semi-privately looking
at the coin.}
\label{epikfigure}
\end{figure}

@@

It remains to define the Kripke relation on guarded strings from an agent specification
as in Figure \ref{epikfigure}. An an agent specification pairs each bare event with a set of bare events,
and so determines a relation between bare events, call it relation $R_a$ for an
agent $e$.  This determines
a relation $\hat{R_a}$ between decorated events, see \ex{Rhat}.
It is generalized to a relation $\doublehat{R}$ on arbitrary guarded strings by
relational Kleene star, as defined in \ex{Rhathat}. The operation $\cdot$ in the
definition is KAT product, which enforces matching of tests.
The number of terms $n$ is construed as satisfying $n\geq 1$.
{\em What is desired for $n=0$?}

@@

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and regular
relations between strings that was designed for use in computationonal phonology and
morphographemics 
\citep{kaplan1994regular,beesley2003finite}.
Current implementations allow for the
definition of functions with the status of defined operators on regular sets and
relations
\citep{hulden2009foma,karttunen2010update}.
Such definitions are used to define an embedding of epistemic KAT in a
string algebra. The methodology follows Section 2 closely.  Let ${\cal K}$ be an
epistemic algebra as described in Section 2.  A given element of ${\cal K}$ is
represented in the the string algebra by the very same set of strings, i.e. by a set
of strings that have the form of a sequence of bare event symbols, with interleaved
Boolean vectors. Product in the KAT can not be modeled as concatenation in the string
algebra, because this would not enforce identity of states, and would result in
lengthening Boolean vectors at the concatenation point.  Instead, KAT product and KAT
Kleene star are defined operations in the string algebra, see Figure \ref{fstfigure1}.  The
operations concatenate in the string algebra, delete strings with non-matching tests,
and then delete the second of two tests create a well-formed guarded string.

@@

\caption{Translation into Fst of KAT product and KAT Kleene star.}
\label{fstfigure1}
\end{figure}

A given bare event such as $a_1$ (Aly looks at heads) is in the KAT algebra a set of
bare events decorated with compatible tests on each side, semantically $\{ 10 a_1 10
\}$ in this case.  This is a unit set rather than a guarded string, because elements
of the KAT algebra are sets. Worlds in the KAT algebra are defined by sequencing
events using {\em Kst}. The operation enforces compatibility of states, so that
$(a_1 + a_0)(b_1 + b_0)$ contains two worlds rather than four.  The program in
Figure \ref{epikfigure} as interpreted in FST defines a countably infinite set of possible worlds by
KAT Kleene closure as $\mbox{\em Kst}(a_1 + a_0 +b_1 + b_0)$, and an algebra of
propositions as regular sets of strings drawn from this space of worlds.

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for each
agent. This determines a relation in the guarded string algebra a relation between
bare events decorated with compatible tests.  For agent Aly, this is the relation
described in \ex{alyrel1} as a set of ordered pairs.

@@

\caption{Definition in Fst of the Kleene concatenation closure of
  a relation between guarded strings.}
\label{fstfigure2}
\end{figure}

@@

This scheme provides for an interpretation of the language of Epik
terms that is defined in Figure \ref{termfigure}.  An Epik specifications such as
Figure 1 is translated to a straight-line program of the finite state
calculus that defines constants and functions, including the ones from
Figures 2 and 3.  In Xfst or Foma, which are interpreters for the
finite state calculus, the program is read, and then propositional
terms can be mapped to finite state machines that represent sets of
guarded strings, interpreted as propositions.

@@

\caption{The language of Epik terms. Propositional terms are interpreted
  as sets of guarded strings.}
\label{termfigure}
\end{figure}

An architecture of interpretation by transtlation is employed, where English
sentences are mapped to terms in the logical language (--) via an interpreted
grammar, and these terms are in turn interpreted as propositions (sets of possible
worlds). For the latter, there are options of translation into the finite state
calculus in order to represent propositions as finite state machines (Section 3), and
representation in Haskell via lazy lists of guarded strings (Section 4).  The grammar
is a semantically interpreted multi-modal categorial grammar, consisting of a lexicon
of words, their categorial types, and interpretations in a logical lambda language.
Figure \ref{phenomenafigure} lists phenomena that are covered.

\begin{figure}  
\begin{tabular}[t]{l|l}
  Basic statives & It's heads. \\
  & It's tails.\\ \hline
That-complement & Amy knows \\ & \hspace{2mm} that it's heads. \\ \hline
Wh-complement & Amy knows \\ & \hspace{2mm} whether its heads. \\ \hline
Negation & Bob doesn't know \\ & \hspace{2mm} that it isn't heads. \\ \hline 
Tensed and base &  Bob knows \\
verbal forms & \hspace{2mm} that it's heads.      \\ 
\hspace{1mm} & Bob doesn't know \\ & \hspace{2mm} that it's heads.\\ \hline
Sentence conjunction & It's heads and Bob \\ & \hspace{2mm} doesn't know \\ & \hspace{2mm} that it's heads \\ \hline
Predicate conjunction & Bob knows that \\ & Amy knows whether \\ & it's heads and
\\ & doesn't know that \\ & Amy knows that \\ & it's heads.
\end{tabular}
\caption{Phenomena covered in the English grammar fragment.}
\label{phenomenafigure}
\end{figure}

@@

Figure \ref{lexiconfigure} gives illustrative lexical entries. The grammar and semantics are in
certain way optimized for a simple fragment of English concerned with clausal
complementation.  The agent names {\em Amy} and {\em Bob} contribute the epistemic
alternative relations for those agents, rather than individuals.  This is possible
because the agents are never arguments of extensional predicates, so what matters
about the agents is their epistemic alternative relations.  The root verb {\em know}
contributes existential modal force.  The complementizers {\em that} and {\em
  whether} are the heads of their dominating clauses, and assemble an alternative
relation, modal force, and proposition contributed by the complement.  These
complementizers introduce the dual via two negations, in order to arrive universal
modal force.  These moves are offered here as a way of constructing a compact
interpreted grammar. They can easily be reformulated in a more comprehensive
interpreted grammar of English.

\begin{figure}  
\begin{tabular}[t]{l|c|c}
Amy & $e$ & ${R}_a$\\
Bob & $e$ & ${R}_b$\\
it & $d$ & $d$ \\
heads & $d \backslash_D t$ & $\lambda x.0^c{\cdot}h$ \\
tails & $d \backslash_D t$ & $\lambda x.0^c{\cdot}!h$ \\
is & $(d \backslash t)\slash(d \backslash_D t)$ & $\lambda P.\lambda x.P x$ \\
knows & $(e \backslash t)\slash_M t$ &  $\lambda p.\lambda R. {\Diamond}R p$\\
that  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
whether  & \multicolumn{2}{l}{$(((e \backslash t)\slash_M t) \backslash (e \backslash t)) \slash t$} \\
& \multicolumn{2}{l}{$\lambda p.\lambda m.\lambda R.{\sim}(m ({\sim}p) R)$} \\
& \multicolumn{2}{r}{$+ {\sim}(m p R)$} \\
\end{tabular}
\caption{Partial categorial grammar lexicon. The first column has a word form.
  the second column a categorial type, and third column a semantic translation in
a logical language that extends the Epik term language with lambda.}
\label{lexiconfigure}
\end{figure}

@@

The semantic translations in the third column of Figure \ref{lexiconfigure} use the Epik term
language, incremented with lambda.  The body of $\lambda x.0^c{\cdot}h$, which is the
semantic lexical entry for {\em heads}, is a term denoting the set of all worlds
where the coin is heads, represented as the set of all guarded strings that end with
a Boolean valuation where the primitive proposition $h$ (it's heads) is true.  There
is $\lambda x$ at the front because of a correspondence the grammar formalism uses a
correspondence between syntactic and semantic types. However, it does not bind
anything, because sentences such as {\em it isn't heads} have an expletive subject.
The body ${\Diamond}R p$ of $\lambda p.\lambda R. {\Diamond}R p$, which is the
semantic lexical entry of {\em knows}, is an Epik term denoting the pre-image of the
set of worlds $p$ according to the relation $\doublehat{R}$ between guarded strings
that is determined by the event-level relation $R$.  This is not the right semantics
for {\em Amy knows that it's heads}, because it is an existential modality ${\Diamond_R}p$,
rather than an universal modality ${\Box_R}p$.  This is corrected by the complementizer
{\em that} or {\em whether}, which introduces the dual.
